How to select predictive models for decision-making or causal inference?

### 2023-12-08 how_to_select_causal_models.tex
adding jamia refs on causal inference

Progress in machine learning brings predictive models to new health data
\cite{beam2018big,rajkomar2019machine}, with the promise of precision medicine.
Predictive models are increaslingly used for image analysis
\cite{khojaste2022deep,zhang2019radiological,yala2019deep,shen2019deep,nassif2022breast},
routinely-collected patient records
\cite{mooney2018bigdata,desai2020comparison,simon2018predicting} or clinical
notes \cite{horng2017creating,wang2020prediction,spasic2020clinical}. Data may
be difficult to control and model, but the accuracy of the prediction can be
verified on left-out data
\cite{altman2009prognosis,poldrack2020establishment,varoquaux2022evaluating}.
Given a model predicting a health outcome, precision medicine would like it to
guide decisions: will an individual benefit from an intervention such as surgery
\cite{fontana2019can}? Contrasting predictions with and without the treatment
gives an answer, but statistical validity requires causal inference
\cite{snowden_implementation_2011,blakely2020reflection}.

@@

Predictive modeling bridges to causal-inference with the so-called \emph{outcome
    models}, or G-computation, G-formula \cite{robins_role_1986}, Q-model
\cite{snowden_implementation_2011}, conditional mean regression
\cite{wendling_comparing_2018}. A central challenge of inference of treatment
effects is that of confounding: spurious associations between treatment
allocation and baseline health, \emph{eg} only prescribing a drug to mild cases
\citep{hernan_causal_2020,vanderweele2019principles}.
% Controlled allocation of treatment, as in randomized controlled trials (RCTs),
% alleviate this concern. Yet most machine-learning models are trained on
% \emph{observational} data, close to real-word practice
% \cite{black1996we,hernan_methods_2021} but challenging for causal inference.
Causal inference has been central to epidemiology, typically with methods that
model treatment assignment \cite{austin_moving_2015,grose_use_2020}, based on
propensity scores \cite{rosenbaum_central_1983}. But outcome modeling go
beyond average effects, estimating individualized or conditional average
treatment effects (CATE), central to precision medicine.
%
For this purpose, such methods are also invaluable on randomized trials
\cite{su2018random,lamont2018identification,hoogland2021tutorial}.

@@

%\idea{There is a variety of models for g-estimation}
% There is an explosion of outcome modeling or machine learning methods, either 
% developed for
% For instance, many deep-learning methods have been developed for medical
% image analysis \cite{shen2017deep,monshi2020deep}.
Outcome-modeling methods, even when specifically designed for causal
inference are numerous: Bayesian Additive Regression Trees
\cite{hill_bayesian_2011}, Targeted Maximum Likelihood Estimation
\cite{laan_targeted_2011,schuler_targeted_2017}, causal boosting
\cite{powers_methods_2018}, causal multivariate adaptive regression
splines \cite{powers_methods_2018}, random forests
\cite{wager_estimation_2018, athey_generalized_2019},
Meta-learners \cite{kunzel_metalearners_2019}, R-learners
\cite{nie_quasioracle_2017}, Doubly robust estimation
\cite{chernozhukov_double_2018}...
%\idea{The variety of estimators calls for model selection procedures}
The wide variety of methods raises the problem
of selecting between different estimators based on the data at hand. %TODO: might be removed since repeated two sentences below
%
Indeed, estimates of treatment effects can vary markedly across different
predictive models (as shown by \cite{dorie_automated_2019} and developed in
Appendix \ref{apd:toy_example:acic_2016_ate_variability}).

>>>>

Progress in machine learning brings predictive models to new health data
\cite{beam2018big,rajkomar2019machine}, with the promise of precision medicine.
Predictive models are increaslingly used for image analysis
\cite{khojaste2022deep,zhang2019radiological,yala2019deep,shen2019deep,nassif2022breast},
routinely-collected patient records
\cite{mooney2018bigdata,desai2020comparison,simon2018predicting} or clinical
notes \cite{horng2017creating,wang2020prediction,spasic2020clinical}. Data may
be difficult to control and model, but the accuracy of the prediction can be
verified on left-out data
\cite{altman2009prognosis,poldrack2020establishment,varoquaux2022evaluating}.
Given a model predicting a health outcome, precision medicine would like it to
guide decisions: will an individual benefit from an intervention such as surgery
\cite{fontana2019can}? Contrasting predictions with and without the treatment
gives an answer, but statistical validity requires causal inference
\cite{snowden_implementation_2011,sperrin2019explicit,blakely2020reflection}.

@@

Predictive modeling bridges to causal-inference with the so-called \emph{outcome
    models}, or G-computation, G-formula \cite{robins_role_1986}, Q-model
\cite{snowden_implementation_2011}, conditional mean regression
\cite{wendling_comparing_2018}. A central challenge of inference of treatment
effects is that of confounding: spurious associations between treatment
allocation and baseline health, \emph{eg} only prescribing a drug to mild cases
\citep{hernan_causal_2020,vanderweele2019principles}.
% Controlled allocation of treatment, as in randomized controlled trials (RCTs),
% alleviate this concern. Yet most machine-learning models are trained on
% \emph{observational} data, close to real-word practice
% \cite{black1996we,hernan_methods_2021} but challenging for causal inference.
Causal inference has been central to epidemiology, typically with methods that
model treatment assignment \cite{austin_moving_2015,casucci2018estimating,grose_use_2020}, based on
propensity scores \cite{rosenbaum_central_1983}. But outcome modeling go
beyond average effects, estimating individualized or conditional average
treatment effects (CATE), central to precision medicine.
%
For this purpose, such methods are also invaluable on randomized trials
\cite{su2018random,lamont2018identification,hoogland2021tutorial}.

@@

%\idea{There is a variety of models for g-estimation}
% There is an explosion of outcome modeling or machine learning methods, either 
% developed for
% For instance, many deep-learning methods have been developed for medical
% image analysis \cite{shen2017deep,monshi2020deep}.
Outcome-modeling methods, even when specifically designed for causal
inference are numerous: Bayesian Additive Regression Trees
\cite{hill_bayesian_2011}, Targeted Maximum Likelihood Estimation
\cite{laan_targeted_2011,schuler_targeted_2017}, causal boosting
\cite{powers_methods_2018}, causal multivariate adaptive regression
splines \cite{powers_methods_2018}, random forests
\cite{wager_estimation_2018, athey_generalized_2019},
Meta-learners \cite{kunzel_metalearners_2019}, R-learners
\cite{nie_quasioracle_2017}, Doubly robust estimation
\cite{chernozhukov_double_2018}...
%\idea{The variety of estimators calls for model selection procedures}
The wide variety of methods raises the problem
of selecting between different estimators based on the data at hand. %TODO: might be removed since repeated two sentences below
%
Indeed, estimates of treatment effects can vary markedly across different
predictive models \cite{fang2019applying,dorie_automated_2019} (illustration in
Appendix \ref{apd:toy_example:acic_2016_ate_variability}).
